\documentclass[aspectratio=169]{beamer}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}

% --- Setup Cooper Union Colors ---
\definecolor{CooperRed}{HTML}{A51C30}
\definecolor{CooperDark}{HTML}{212121}

\setbeamercolor{palette primary}{bg=CooperRed,fg=white}
\setbeamercolor{palette secondary}{bg=CooperRed,fg=white}
\setbeamercolor{palette tertiary}{bg=CooperDark,fg=white}
\setbeamercolor{title in head/foot}{bg=CooperDark,fg=white}
\setbeamercolor{title}{fg=CooperRed}
\setbeamercolor{frametitle}{bg=CooperDark,fg=white}

% --- Title Info ---
\title{ECE 251: Computer Architecture}
\subtitle{Week 11: Pipelining \& Hazard Resolution}
\author{Professor Rob Marano}
\institute{The Cooper Union}
\date{Spring 2026}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{1. Transition to True Hardware Pipelining}
\begin{itemize}
    \item \textbf{Core Concept}: Multiple instructions are actively executed simultaneously across discrete hardware zones (Overlap Execution).
    \item \textbf{The Goal}: Process $N$ instructions in $N$ clock cycles, driving Cycles Per Instruction ($CPI$) down to $1.0$.
    \item \textbf{The Reality}: The global machine clock is constrained strictly by the slowest hardware stage (e.g., Memory retrieval).
    \item \textbf{The Trade-off}: Throughput dramatically increases over Single-Cycle, but the physical duration applied to an individual isolated instruction (Latency) slightly increases due to pipeline logic overhead.
\end{itemize}
\end{frame}

\begin{frame}{2. The Amdahl's Law Limit}
\begin{itemize}
    \item Pipelining overlaps execution, scaling speedup proportionally by hardware stages $k$.
    \item \textbf{Amdahl's Formula}: $\text{Speedup} = \frac{1}{(1 - p) + \frac{p}{k}}$
    \item \textbf{MIPS Assembly Mapping ($p$)}: If $p = 0.8$, $80\%$ of instructions execute flawlessly in parallel (e.g. bypassed ALU math, safely predicted branches).
    \item \textbf{The Penalty Mapping ($1-p$)}: $20\%$ of execution triggers a hard structural stop via \textbf{Hazards} (e.g., \texttt{lw} memory stalls, pipeline flushes).
    \item \textbf{The Mathematical Ceiling}: Even as pipelining stages approach infinity ($k \to \infty$), the maximum physical speedup strictly caps at $\frac{1}{1-p}$. Hazards physically dictate CPU limits!
\end{itemize}
\end{frame}

\begin{frame}{2b. Resolving Break-Even Parallelization}
\begin{itemize}
    \item \textbf{Problem}: An 8-stage pipeline ($k=8$) \textit{must} achieve a $4.0\times$ speedup over a single-cycle implementation. Calculate the fraction of instructions ($p$) that must be completely free of hazards to hit this target.
    \item \textbf{Algebraic Execution}:
    \begin{align*}
        4.0 &= \frac{1}{(1-p) + \frac{p}{8}} \\
        (1-p) + \frac{p}{8} &= 0.25 \\
        1 - \frac{7p}{8} &= 0.25 \\
        \frac{7p}{8} &= 0.75 \implies 7p = 6 \implies p \approx 85.71\%
    \end{align*}
    \item \textbf{Result}: The compiler and hardware forwarders must orchestrate a flawless bypass environment for $85.71\%$ of executed code.
\end{itemize}
\end{frame}

\begin{frame}{3. Establishing Pipeline Registers}
Because data must be preserved securely while instructions shift into the next functional stage, we partition the datapath using large $D$-Flip-Flop arrays called \textbf{Pipeline Registers}.
\vspace{0.2cm}
\begin{enumerate}
    \item \textbf{IF/ID}: Holds fetched 32-bit instruction payload and sequential PC.
    \item \textbf{ID/EX}: Holds Registry data ($A$, $B$), Sign-Extended Immediate, and target Register Addresses.
    \item \textbf{EX/MEM}: Holds computed ALU output and target Write Destination address.
    \item \textbf{MEM/WB}: Holds loaded RAM payload preparing for destination registry push.
\end{enumerate}
\end{frame}

\begin{frame}[fragile]{4. SystemVerilog: The Pipeline Register Stage}
\begin{minted}[fontsize=\scriptsize,bgcolor=lightgray]{systemverilog}
// SystemVerilog ID/EX Pipeline Register DFF
module id_ex_reg (input  logic        clk, reset,
                  input  logic        RegWriteD, MemWriteD, // Control
                  input  logic [31:0] RD1D, RD2D, ImmD,     // Data
                  input  logic [4:0]  RsD, RtD, RdD,        // Targeting
                  output logic        RegWriteE, MemWriteE, 
                  output logic [31:0] RD1E, RD2E, ImmE,
                  output logic [4:0]  RsE, RtE, RdE);

  always_ff @(posedge clk or posedge reset) begin
    if (reset) begin
      RegWriteE <= 0; // Evaluate No-Op flush...
    end else begin
      RegWriteE <= RegWriteD;
      RD1E      <= RD1D;
      RsE       <= RsD;       // Locked for Hazard evaluation downstream
    end
  end
endmodule
\end{minted}
\end{frame}

\begin{frame}{5. Data Hazards: RAW (Read-After-Write)}
\begin{itemize}
    \item \textbf{Data Hazards} manifest when a physical target instruction attempts to consume algebraic variables generated by an older instruction still evaluating inside the pipeline.
    \item Example (Read-After-Write RAW calculation bypass):
    \begin{itemize}
        \item \texttt{add \$s0, \$t0, \$t1} $\to$ \texttt{\$s0} locked for modification via ALU
        \item \texttt{sub \$t2, \$s0, \$t3} $\to$ Immediately demands \texttt{\$s0} for subtraction input
    \end{itemize}
    \item \textbf{Solution}: We do not wait for the fifth stage (WB). The ALU calculates \texttt{\$s0} locally in stage 3 (EX). We algorithmically \textbf{Forward} (bypass) the data natively backwards utilizing multiplexers.
    \item \textbf{Physical Stalling (The Bubble)}: If the data isn't algebraically ready (e.g. a memory load \texttt{lw} taking 4 stages), forwarding is physically impossible in time. We must stall the pipeline by synthesizing an inert algorithmic bubble! \textit{Did you know? MIPS architecture natively invokes the raw code} \texttt{0x00000000} \textit{translated identically as} \texttt{sll \$zero, \$zero, 0} \textit{to safely delay execution 1 cycle without corrupting registries!}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{6. SystemVerilog: The Hazard / Forwarding Unit}
\begin{minted}[fontsize=\scriptsize,bgcolor=lightgray]{systemverilog}
module hazard_unit (input  logic [4:0] RsE, RtE,   // Current Stage Src Regs
                    input  logic [4:0] WriteRegM,  // Prev Stage Dest Reg
                    input  logic       RegWriteM,  // Validation Flag
                    output logic [1:0] ForwardAE, ForwardBE);

  // Forwarding evaluation for ALU Source Port A
  always_comb begin
    if ((RegWriteM) && (WriteRegM != 0) && (WriteRegM == RsE)) begin
        // The execution source register mathematically collides with 
        // the pending write destination of the previous cycle.
        ForwardAE = 2'b10; // Trigger MUX to route ALUOutM backwards to ALU Input A
    end else begin
        ForwardAE = 2'b00; // Use clean baseline Register File extraction
    end
  end
endmodule
\end{minted}
\end{frame}

\begin{frame}{7. Control Hazards: Pipeline Penalties}
\begin{itemize}
    \item \textbf{The Problem}: The processor blindly fetches sequential code (\texttt{PC + 4}). Upon striking a branch instruction (\texttt{beq}), the true execution path is unknown until the ALU evaluates conditions inside the third EX phase.
    \item \textbf{Software Fix (See MIPS Run - Branch Delay Slot)}: Early architectures conserved transistor budgets by deliberately exposing pipelines to compilers. The hardware mechanically executed the instruction directly following the branch. Software C-Compilers had to meticulously pack this ``Delay Slot'' with safe, independent execution logic or risk corrupted state vectors.
    \item \textbf{Hardware Evolution (Predict \& Flush)}: By predicting \textit{Branch Not Taken}, the datapath optimistically marches forward. If evaluated incorrectly as \textit{Taken}, the pipeline physically halts, triggers the \texttt{reset} line on \texttt{IF/ID} bounds, and structurally collapses the rogue payload into harmless \texttt{0's} (NOP bubbles).
\end{itemize}
\end{frame}

\begin{frame}{8. Exceptions \& Interrupts}
\begin{itemize}
    \item Pipelining inherently disrupts orderly procedural progression. A division-by-zero execution fault or hardware interrupt must freeze standard datapath fetching operations immediately.
    \item The logic traps the specific active \texttt{PC} sequence inside the \texttt{EPC} (Exception Program Counter) tracking register, explicitly squashes pipeline operations locally, and drives processing towards the local OS Exception Handler boot protocol vector mapping mapping.
\end{itemize}
\end{frame}

\begin{frame}{9. Week 11 Summary}
\begin{itemize}
    \item \textbf{Pipeline Strategy}: Divide datapath execution linearly over discrete logical $D$-flip-flop boundary constraints natively driving isolated logic stages concurrently over absolute clock phases.
    \item \textbf{Hazard Physics}: Resolving algebraic resource overlaps necessitates explicit mathematical handling logic physically designed across tracking registers vs core ALU forwarding constraints.
    \item \textbf{Amdahl's Cap}: Hazards define the sequential ceiling $1-p$ capping overall pipeline scaling potential.
    \item \textbf{Next Session}: Extending into Advanced Interrupts and Memory Hierarchy evaluations defining system-level architectural caching logic frameworks.
\end{itemize}
\end{frame}

\end{document}
